---
title: "S1 Appendix"
author: 
  - Anna Niehues^[Radboud University Medical Center, Anna.Niehues@radboudumc.nl]
output: bookdown::html_document2
bibliography: bibliography.bib
---

## Results of gene set enrichment analyses

Gene-set enrichment analyses (GSEA) were performed using the R/Bioconductor package fgsea \cite{https://doi.org/10.1101/060012} and gene sets retrieved from the Reactome Pathway Database \cite{https://doi.org/10.1093/nar/gkz1031}. Genes were ranked by $-\log_{10}(p_{b})*|\beta_{b}|$ with $p_{b} $ = bacon-corrected p-value and $beta_{b}$ = bacon-corrected effect size. The number of permutations for initial estimation of p-values was set to \num{1e4}; the boundary for calculating p-values was set to \num{1e-50}.


```{r include=FALSE}
plot_pathways <- function(GSEA_by_comparison_by_cohort, 
                          outfile = NULL,
                          alpha = 0.05,
                          max_pathways = FALSE) {
  # data frame with pathways from all GSEAs, i.e. for multiple models 
  gsea_merged <- data.table::rbindlist(
    lapply(names(GSEA_by_comparison_by_cohort), function(x) {
      df <- GSEA_by_comparison_by_cohort[[x]]$fgsea
      df$model <- x
      df}))
  # convert pathway column to factor for ordered plots
  require(dplyr)
  merge_for_factor <- gsea_merged %>%
    group_by(pathway) %>%
    summarize(order = sum(abs(NES)*-log10(padj)))
  gsea_merged$pathway <- factor(
    gsea_merged$pathway, 
    levels = merge_for_factor$pathway[
      order(merge_for_factor$order, decreasing = FALSE)])
  # union of significant pathways
  pathway_signif_union <- Reduce(
    union, 
    lapply(GSEA_by_comparison_by_cohort, function(x) {
      x$fgsea[x$fgsea$padj < alpha,]$pathway}))
  #print(length(pathway_signif_union))
  # if too many pathways, decrease alpha for plot
  if (max_pathways) {
    tmp_alpha <- alpha
    while (length(pathway_signif_union) > max_pathways) {
      tmp_alpha <- tmp_alpha / 10
      pathway_signif_union <- Reduce(
        union,
        lapply(GSEA_by_comparison_by_cohort, function(x) {
          x$fgsea[x$fgsea$padj < tmp_alpha,]$pathway}))
    }
  }
  # plot pathways, normalized enrichment scores and adjusted p-values
  require(ggplot2)
  if (is.null(outfile)) {
    print("No significantly enriched pathways.")
  } else {
    tmp <- gsea_merged[order(gsea_merged$pathway, decreasing = TRUE), ][1:20]
    tmp$leadingEdge <- vapply(tmp$leadingEdge, paste, collapse = ", ", character(1L))
    write.csv(tmp, file = outfile)
  }
  if (length(pathway_signif_union) > 0) {
    ggplot(gsea_merged[gsea_merged$pathway %in% pathway_signif_union, ]) +
      geom_segment(aes(y = pathway, x = NES, xend = 0, yend = pathway, )) +
      geom_point(aes(y = pathway, x = NES, 
                     fill = ifelse(padj < alpha, -log10(padj), NA), 
                     size = size), shape = 21) +
      theme(axis.text.y = element_text(size = 6)) +
      scale_fill_viridis_c(na.value = "white") +
      facet_grid(cols = vars(model)) 
  }
}
```

## eGFR

```{r, fig.width = 12, fig.height = 43, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome eGFR (s_low_eGFR, metabolic surrogate).", echo=F}
plot_pathways(readd(meta_GSEA_by_comparison_eGFR),
              outfile = file.path(readd(outdir), "GSEA_eGFR.csv"))
```

## Triglycerides

```{r, fig.width = 12, fig.height = 38, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome triglycerides (s_high_triglycerides, metabolic surrogate; triglycerides, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Triglycerides),
              outfile = file.path(readd(outdir), "GSEA_Triglycerides.csv"))
```

## LDL cholesterol

```{r, fig.width = 12, fig.height = 7, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome LDL cholesterol (s_high_ldl_chol, metabolic surrogate; ldlchol, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_LDL.cholesterol),
              outfile = file.path(readd(outdir), "GSEA_LDL cholesterol.csv"))
```

## Total cholesterol

```{r, fig.width = 12, fig.height = 35, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome total cholesterol (s_high_totchol, metabolic surrogate; totchol, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Total.cholesterol),
              outfile = file.path(readd(outdir), "GSEA_Total cholesterol.csv"))
```

## HDL cholesterol

```{r, fig.width = 12, fig.height = 38, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome HDL cholesterol (s_low_hdlchol, metabolic surrogate; hdlchol, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_HDL.cholesterol),
              outfile = file.path(readd(outdir), "GSEA_HDL cholesterol.csv"))
```

## Diabetes

```{r, fig.width = 12, fig.height = 22, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome diabetes (s_diabetes, metabolic surrogate).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Diabetes),
              outfile = file.path(readd(outdir), "GSEA_Diabetes.csv"))
```

## Metabolic syndrome

```{r, fig.width = 12, fig.height = 36, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome metabolic syndrome (s_metabolic_syndrome, metabolic surrogate).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Metabolic.syndrome),
              outfile = file.path(readd(outdir), "GSEA_Metabolic syndrome.csv"))
```

## Sex (male)

```{r, fig.width = 12, fig.height = 5}
plot_pathways(readd(meta_GSEA_by_comparison_Sex..male_))
```

```{r, echo = FALSE, include=FALSE}
# plot_pathways(readd(GSEA_by_comparison_by_cohort_Sex..male._NTR))
# plot_pathways(readd(GSEA_by_comparison_by_cohort_Sex..male._LL))
# plot_pathways(readd(GSEA_by_comparison_by_cohort_Sex..male._LLS))
# plot_pathways(readd(GSEA_by_comparison_by_cohort_Sex..male._RS))
```

## Lipid medication

```{r, fig.width = 12, fig.height = 33, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome lipid medication (s_lipidmed, metabolic surrogate; lipidmed_statins, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Lipid.medication),
              outfile = file.path(readd(outdir), "GSEA_Lipid medication.csv"))
```

## hsCRP

```{r, fig.width = 12, fig.height = 35, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome hsCRP (s_high_hscrp, metabolic surrogate; hscrp, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_hsCRP),
              outfile = file.path(readd(outdir), "GSEA_hsCRP.csv"))
```

## Blood pressure lowering medication

```{r, fig.width = 12, fig.height = 33, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome blood pressure lowering medication (s_blood_pressure_lowering_med, metabolic surrogate).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Blood.pressure.lowering.medication),
              outfile = file.path(readd(outdir), "GSEA_Blood pressure lowering medication.csv"))
```

## Age

```{r, fig.width = 12, fig.height = 29, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome age (s_high_age, metabolic surrogate; sampling_age, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_.High..age),
              outfile = file.path(readd(outdir), "GSEA_(High) age.csv"))
```

## BMI, obesity

```{r, fig.width = 12, fig.height = 31, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome BMI/obesity (s_obesity, metabolic surrogate; bmi, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_BMI..obesity),
              outfile = file.path(readd(outdir), "GSEA_BMI, obesity.csv"))
```


## Hemoglobin

```{r, fig.width = 12, fig.height = 57, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome hemoglobin (s_low_hgb, metabolic surrogate; hgb, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Hemoglobin),
              outfile = file.path(readd(outdir), "GSEA_Hemoglobin.csv"))
```

## White blood cells

```{r, fig.width = 12, fig.height = 37, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome white blood cells (s_low_wbc, metabolic surrogate; wbc, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_White.blood.cells),
              outfile = file.path(readd(outdir), "GSEA_White blood cells.csv"))
```

## Current smoking 

```{r, fig.width = 12, fig.height = 10, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome current smoking (s_current_smoking, metabolic surrogate; smoking_current, reported variable).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Current.smoking),
              outfile = file.path(readd(outdir), "GSEA_Current smoking.csv"))

```

## Alcohol consumption 

```{r, fig.width = 12, fig.height = 21, fig.cap = "GSEA results of meta-analyzed TWAS results for outcome alcohol consumption (s_alcohol_consumption, metabolic surrogate).", echo = FALSE}
plot_pathways(readd(meta_GSEA_by_comparison_Alcohol.consumption),
              outfile = file.path(readd(outdir), "GSEA_Alcohol consumption.csv"))
```

