---
title: "Results of Gene Set Enrichment Analysis (GSEA)"
author: 
  - Anna Niehues^[Radboud University Medical Center, Anna.Niehues@radboudumc.nl]
output: bookdown::html_document2
bibliography: bibliography.bib
---

```{r}
library(drake)
loadd(epismoker_variables)
loadd(surrogates_variables)
get_covariatetype <- function(outcome_variable) {
  if (outcome_variable %in% epismoker_variables) {
    covariate_type <- "Epigenetic surrogate"
  } else if (outcome_variable %in% surrogates_variables) {
    covariate_type <- "Metabolic surrogate"
  } else {
    covariate_type <- "Reported variable"
  }
  return(covariate_type)
}
```

```{r gsea_data}
require(dplyr)
rows = list()
row_idx = 0
# rows_meta = list()
# row_idx_meta = 0
for (comparison in models_json) {
  # if (endsWith(names(comparison$models)[[1]], "_z")) {
    # print(comparison$comparison)
    # GSEA of meta-analyzed TWAS
    meta_gsea_name <- paste0("meta_GSEA_by_comparison_", comparison$comparison)
    meta_gsea_name <- gsub("[, ()]", ".", meta_gsea_name)
    if (endsWith(meta_gsea_name, ".")) {
      substr(meta_gsea_name, nchar(meta_gsea_name), nchar(meta_gsea_name)) <- "_"
    }
    meta_gsea_results <- readd(meta_gsea_name, character_only = TRUE)
    save(meta_gsea_results,
         file = file.path(readd(outdir), 
                          comparison$comparison, 
                          paste0(meta_gsea_name, ".RData", sep = "")))
    pathway_intersect <- Reduce(intersect, lapply(meta_gsea_results, function(x) {
      x$fgsea[x$fgsea$padj < 0.05,]$pathway}))
    # print(length(pathway_intersect))
    row_idx <- row_idx + 1
    rows[[row_idx]] <- list(
      comparison = comparison$comparison,
      cohort = "Meta-analysis",
      model = "Intersection",
      # scaling = "z",
      covariate_type = "Intersection",
      num_pathways = length(pathway_intersect)
    )
    for (model in names(meta_gsea_results)) {
      if (endsWith(model, "_z")) {
        model_base_name <- substr(model, 1, nchar(model)-2)
        # scaling <- "z"
      } else {
        model_base_name <- model
        # scaling <- "none"
      }
      covariate_type <- get_covariatetype(model_base_name)
      if (covariate_type != "Epigenetic surrogate") {
        row_idx <- row_idx + 1
        rows[[row_idx]] <- list(
          comparison = comparison$comparison,
          cohort = "Meta-analysis",
          model = model,
          # scaling = scaling,
          covariate_type = covariate_type,
          num_pathways = length(meta_gsea_results[[model]]$fgsea[
            meta_gsea_results[[model]]$fgsea$padj < 0.05,]$pathway)
        )
      }
    }
    # individual cohorts GSEA
    for (cohort in comparison$cohorts) {
      gsea_name <- paste0(
        "GSEA_by_comparison_by_cohort_", comparison$comparison, "_", cohort)
      gsea_name <- gsub("[, ()]", ".", gsea_name)
      gsea_results <- readd(gsea_name, character_only = TRUE)
      # print(cohort)
      # lapply(gsea_results, function(x) {print(sum(x$padj < 0.05))})
      pathway_intersect <- Reduce(intersect, lapply(gsea_results, function(x) {
        x$fgsea[x$fgsea$padj < 0.05,]$pathway}))
      # print(length(pathway_intersect))
      row_idx <- row_idx + 1
      rows[[row_idx]] <- list(
        comparison = comparison$comparison,
        cohort = cohort,
        model = "Intersection",
        # scaling = scaling,
        covariate_type = "Intersection",
        num_pathways = length(pathway_intersect)
      )
      for (model in names(gsea_results)) {
        if (endsWith(model, "_z")) {
          model_base_name <- substr(model, 1, nchar(model)-2)
          # scaling <- "z"
        } else {
          model_base_name <- model
          # scaling <- "none"
        }
        covariate_type <- get_covariatetype(model_base_name)
        if (covariate_type != "Epigenetic surrogate") {
          row_idx <- row_idx + 1
          rows[[row_idx]] <- list(
            comparison = comparison$comparison,
            cohort = cohort,
            model = model,
            # scaling = scaling,
            covariate_type = covariate_type,
            num_pathways = length(gsea_results[[model]]$fgsea[
              gsea_results[[model]]$fgsea$padj < 0.05,]$pathway)
          )
        }
      }
    }
  # }
}
gsea_df <- data.table::rbindlist(rows)
gsea_df_comparisonlabels <- unique(gsea_df$comparison)
gsea_df_comparisonlabels[gsea_df_comparisonlabels == "Blood pressure lowering medication"] <- "Blood pressure\nlowering medication"
gsea_df$Comparison <- factor(
  gsea_df$comparison,
  levels = unique(gsea_df$comparison),
  labels = gsea_df_comparisonlabels)
gsea_df$`Variable type` <- factor(gsea_df$covariate_type,
                                  levels = c("Reported variable",
                                             "Metabolic surrogate",
                                             "Epigenetic surrogate",
                                             "Intersection"),
                                  labels = c("Reported \nvariable",
                                             "Metabolic\nsurrogate",
                                             "Epigenetic\nsurrogate",
                                             "Intersection"))
gsea_df$`Cohort` <- factor(gsea_df$cohort,
                                  levels = c("Meta-analysis", "LL", "LLS", "NTR", "RS"),
                                  labels = c("Meta-\nanalysis", "LL", "LLS", "NTR", "RS"))


```


```{r gseanumpathways, fig.cap = "Number of pathways found in in gene set enrichment analysis (adjusted p-value < 0.05)"}
require(ggplot2)
# http://jfly.iam.u-tokyo.ac.jp/color/
cbf_palette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
ggp_gseapercohort <- ggplot(gsea_df) +
  geom_bar(aes(x = `Variable type`, y = num_pathways, fill = `Variable type`), 
           stat = "identity") +
  facet_grid(
    cols = vars(Comparison),
    rows = vars(Cohort)) +
    #switch = "x") +
  ylab("Number of pathways") +
  # scale_fill_brewer(palette = "Dark2") + 
  scale_fill_manual(values = cbf_palette) +
  theme_light(base_size = 6) +
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x= element_blank(),
    #axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text.x = element_text(angle = 90),
    strip.background =element_rect(fill = "lightgray"),
    strip.text = element_text(colour = "black")) 

ggsave(file.path(readd(outdir), "fig5.pdf"), ggp_gseapercohort, 
       width = 12, height = 7, units = "cm", dpi = 1200)
write.csv(gsea_df,
     file = file.path(readd(outdir), "fig5_data.csv"))
ggp_gseapercohort

```


```{r pathwayplotfunc}
plot_pathways <- function(GSEA_by_comparison_by_cohort, 
                          alpha = 0.05,
                          max_pathways = FALSE) {
  require(fgsea)
  lapply(GSEA_by_comparison_by_cohort, function(x) {
    topPathwaysUp <- x$fgsea[ES > 0][head(order(pval), n=10), pathway]
    topPathwaysDown <- x$fgsea[ES < 0][head(order(pval), n=10), pathway]
    topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
    plot.new()
    plotGseaTable(x$pathways[topPathways], x$ranks, x$fgsea, 
                  gseaParam = 0.5)
  })
  # data frame with pathways from all GSEAs, i.e. for multiple models 
  gsea_merged <- data.table::rbindlist(
    lapply(names(GSEA_by_comparison_by_cohort), function(x) {
      df <- GSEA_by_comparison_by_cohort[[x]]$fgsea
      df$model <- x
      df}))
  # save(gsea_merged, file = "GSEAexample.RData")
  # convert pathway column to factor for ordered plots
  require(dplyr)
  merge_for_factor <- gsea_merged %>%
    group_by(pathway) %>%
    summarize(order = sum(abs(NES)*-log10(padj)))
  gsea_merged$pathway <- factor(
    gsea_merged$pathway, 
    levels = merge_for_factor$pathway[
      order(merge_for_factor$order, decreasing = FALSE)])
  # union of significant pathways
  pathway_signif_union <- Reduce(
    union, 
    lapply(GSEA_by_comparison_by_cohort, function(x) {
      x$fgsea[x$fgsea$padj < alpha,]$pathway}))
  # if too many pathways, decrease alpha for plot
  if (max_pathways) {
    tmp_alpha <- alpha
    while (length(pathway_signif_union) > max_pathways) {
      tmp_alpha <- tmp_alpha / 10
      pathway_signif_union <- Reduce(
        union,
        lapply(GSEA_by_comparison_by_cohort, function(x) {
          x$fgsea[x$fgsea$padj < tmp_alpha,]$pathway}))
    }
  }
  # plot pathways, normalized enrichment scores and adjusted p-values
  require(ggplot2)
  ggplot(gsea_merged[gsea_merged$pathway %in% pathway_signif_union, ]) +
    geom_segment(aes(y = pathway, x = NES, xend = 0, yend = pathway, )) +
    geom_point(aes(y = pathway, x = NES, 
                   fill = ifelse(padj < alpha, -log10(padj), NA), 
                   size = size), shape = 21) +
    theme(axis.text.y = element_text(size = 6)) +
    scale_fill_viridis_c(na.value="white") +
    facet_grid(cols = vars(model)) 
}
```

```{r}
plot_enrichment <- function(gsea_results, pathway) {
  require(fgsea)
  single_plots <- lapply(gsea_results, function(x) {
    plotEnrichment(x$pathways[[pathway]], x$ranks)})
  cowplot::plot_grid(
    single_plots[[1]],
    single_plots[[2]],
    labels = names(gsea_results)
  )
}
```

## eGFR

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_eGFR), max_pathways = 100)
```

## Triglycerides

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_Triglycerides), max_pathways = 100)
```

```{r}
plot_enrichment(readd(meta_GSEA_by_comparison_Triglycerides), "Translation")
plot_enrichment(readd(meta_GSEA_by_comparison_Triglycerides), "Cap-dependent Translation Initiation")
```

## LDL cholesterol

```{r, fig.width = 10, fig.height = 9}
plot_pathways(readd(meta_GSEA_by_comparison_LDL.cholesterol))
```

```{r}
plot_enrichment(readd(meta_GSEA_by_comparison_LDL.cholesterol), "Cholesterol biosynthesis")
plot_enrichment(readd(meta_GSEA_by_comparison_LDL.cholesterol), "Signal Transduction")
```

## Total cholesterol

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_Total.cholesterol), max_pathways = 100)
```

```{r}
plot_enrichment(readd(meta_GSEA_by_comparison_Total.cholesterol), "Immune System")
plot_enrichment(readd(meta_GSEA_by_comparison_Total.cholesterol), "Innate Immune System")
```

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(GSEA_by_comparison_by_cohort_Total.cholesterol_NTR), max_pathways = 100)
plot_pathways(readd(GSEA_by_comparison_by_cohort_Total.cholesterol_RS), max_pathways = 100)
```

## HDL cholesterol

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_HDL.cholesterol), max_pathways = 100)
```

```{r}
plot_enrichment(readd(meta_GSEA_by_comparison_HDL.cholesterol), "Translation")
```

## Diabetes

```{r, fig.width = 10, fig.height = 19}
plot_pathways(readd(meta_GSEA_by_comparison_Diabetes))
```

## Metabolic syndrome

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_Metabolic.syndrome), max_pathways = 100)
```

## Sex (male)

```{r, fig.width = 10, fig.height = 7}
plot_pathways(readd(GSEA_by_comparison_by_cohort_Sex..male._NTR))
```


## Lipid medication

```{r, fig.width = 10, fig.height = 7}
plot_pathways(readd(meta_GSEA_by_comparison_Lipid.medication), max_pathways = 50)
```

## hsCRP

More significantly enriched pathways for meta-analysed TWASs in case of metabolic surrogates



```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_hsCRP), max_pathways = 100)
```

```{r}
plot_enrichment(readd(meta_GSEA_by_comparison_hsCRP), "Translation")
plot_enrichment(readd(meta_GSEA_by_comparison_hsCRP), "Metabolism")
```

## Blood pressure lowering medication

```{r, fig.width = 10, fig.height = 7}
plot_pathways(readd(meta_GSEA_by_comparison_Blood.pressure.lowering.medication), max_pathways = 50)
```

## Age

```{r, fig.width = 10, fig.height = 7}
plot_pathways(readd(meta_GSEA_by_comparison_.High..age), max_pathways = 50)
```

```{r}
plot_enrichment(readd(meta_GSEA_by_comparison_.High..age), 
                "Respiratory electron transport, ATP synthesis by chemiosmotic coupling, and heat production by uncoupling proteins.")
plot_enrichment(readd(meta_GSEA_by_comparison_.High..age), "Innate Immune System")
plot_enrichment(readd(meta_GSEA_by_comparison_.High..age), "Neutrophil degranulation")
```

```{r, fig.width = 10, fig.height = 7}
plot_pathways(readd(GSEA_by_comparison_by_cohort_.High..age_LLS), max_pathways = 50)
```

Discussion: https://jlb.onlinelibrary.wiley.com/doi/pdf/10.1002/jlb.67.1.40 describe age-related decline of neutrophil function, adn possible association with increased susceptibility to infections




## BMI, obesity

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_BMI..obesity), max_pathways = 100)
```

```{r}
plot_enrichment(readd(meta_GSEA_by_comparison_BMI..obesity), "Translation")
plot_enrichment(readd(meta_GSEA_by_comparison_BMI..obesity), "Signal Transduction")
plot_enrichment(readd(meta_GSEA_by_comparison_BMI..obesity), "Response of EIF2AK4 (GCN2) to amino acid deficiency")
plot_enrichment(readd(meta_GSEA_by_comparison_BMI..obesity), "Selenoamino acid metabolism")
plot_enrichment(readd(meta_GSEA_by_comparison_BMI..obesity), "NR1H2 and NR1H3-mediated signaling")
```

## Hemoglobin

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_Hemoglobin), max_pathways = 100)
```

## White blood cells

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_White.blood.cells), max_pathways = 100)
```

## Current smoking 

Epigenetic vs. metabolic score for current smoking.

```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_Current.smoking))
```

```{r}
plot_enrichment(readd(meta_GSEA_by_comparison_Current.smoking), "GTP hydrolysis and joining of the 60S ribosomal subunit")
plot_enrichment(readd(meta_GSEA_by_comparison_Current.smoking), "Translation")
plot_enrichment(readd(meta_GSEA_by_comparison_Current.smoking), "Signal Transduction")
```

## Alcohol consumption 


```{r, fig.width = 10, fig.height = 12}
plot_pathways(readd(meta_GSEA_by_comparison_Alcohol.consumption), max_pathways = 100)
```







